Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-03-11 appendices/appendixa.tex
B-W figures for appendices
fixing floats
changes for printout
cover for new, thicker version

\chapter{Tutorial on setting up time-heterogeneous Epoch Models of substitution in BEAST.\label{app:epoch}}
\chaptermark{Epoch Models of substitution in BEAST}

@@

In this tutorial we will show how to setup, run and analyse data with the epoch models described in \cite{Bielejec2014a}.
We present an artificial example and describe the necessary steps to get an actual analysis up and running.
Example files for this tutorial can be found here \url{http://rega.kuleuven.be/cev/ecv/tutorials/epochExample.zip}.

@@

This tutorial assumes that the user has a basic knowledge of the BEAST framework, knows how to generate an XML input file and can interpret basic output information of an MCMC run.
BEAST runnables for all major platforms can be found here \url{http://beast.bio.ed.ac.uk/}.
You will need a Java runtime environment version 1.5 or greater to run the BEAST executable.
Due to the computational intensity of the computations involved in the epoch model, the BEAGLE library also needs to be installed and used with the BEAST runs discussed here.
See \url{https://code.google.com/p/beagle-lib/w/list} for details on how to install and use the BEAGLE library.

@@

The data in this tutorial was artificially generated by evolving it on an MCC tree generated from a Bayesian phylogenetic analysis of human influenza A hemagglutinin gene sequences sampled through different epidemic seasons \citep{Drummond2010}.
This tree is visualised in Figure~\ref{fig:threeEpoch}.

@@

% FB: Philippe could you work your magic on this figure and mark 3 epochs with transitions at 7 and 15 starting at the tips much like you did for the paper?
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{threeEpoch} 
\caption{
{ \footnotesize 
{\bf Tree topology for time-heterogenous analysis.}
} % END: footnotesize
}
\label{fig:threeEpoch}
\end{figure}

@@

\begin{figure}[H]
\centering
% Tie Fighter plots!
\includegraphics[scale=0.5]{results} 
\caption{
{ \footnotesize 
{\bf 95\% Credible intervals for $\kappa$ parameters.} Horizontal lines mark the true value, black dots indicate posterior mean values.
} % END: footnotesize
}
\label{fig:results}
\end{figure}

>>>>

\chapter{Tutorial on setting up time-heterogeneous Epoch Models of substitution in BEAST.\label{app:epoch}}
\chaptermark{Tutorial: Epoch Models of substitution in BEAST}

@@

In this tutorial we will show how to setup, run and analyse data with the epoch models described in \cite{Bielejec2014a}.
We present an artificial example and describe the necessary steps to get an actual analysis up and running.
Example files for this tutorial can be found at: \\
\url{http://rega.kuleuven.be/cev/ecv/tutorials/epochExample.zip}.

@@

This tutorial assumes that the user has a basic knowledge of the BEAST framework, knows how to generate an XML input file and can interpret basic output information of an MCMC run.
BEAST runnables for all major platforms can be found here \url{http://beast.bio.ed.ac.uk/}.
You will need a Java runtime environment version 1.5 or greater to run the BEAST executable.
Due to the computational intensity of the computations involved in the epoch model, the BEAGLE library also needs to be installed and used with the BEAST runs discussed here.
See:
\begin{center}
\url{https://code.google.com/p/beagle-lib/w/list}
\par\end{center}
\noindent
for details on how to install and use the BEAGLE library.

@@

The data for this tutorial has been artificially generated by evolving it on an MCC tree generated from a Bayesian phylogenetic analysis of human influenza A hemagglutinin gene sequences sampled through different epidemic seasons \citep{Drummond2010}.
This tree is visualised in Figure~\ref{fig:threeEpoch}.

@@

% FB: Philippe could you work your magic on this figure and mark 3 epochs with transitions at 7 and 15 starting at the tips much like you did for the paper?
\begin{figure}[h!]
\centering
\includegraphics[scale=0.35]{threeEpoch} 
\caption{
{ \footnotesize 
{\bf Tree topology for time-heterogenous analysis.}
} % END: footnotesize
}
\label{fig:threeEpoch}
\end{figure}

@@

\begin{figure}[h!]
\centering
% Tie Fighter plots!
\includegraphics[scale=0.5]{results} 
\caption{
{ \footnotesize 
{\bf 95\% Credible intervals for $\kappa$ parameters.} Horizontal lines mark the true value, black dots indicate posterior mean values.
} % END: footnotesize
}
\label{fig:results}
\end{figure}

### 2015-03-11 appendices/appendixb.tex
fixing floats
bib and figures
changes for printout
cover for new, thicker version

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig01}
\caption{
{ \footnotesize 
{\bf Loading MCC tree file.}
} % END: footnotesize
}
\label{fig:01}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig02}
\caption{
{ \footnotesize 
{\bf Message in Terminal.}
} % END: footnotesize
}
\label{fig:02}
\par\end{centering}
\end{figure}

@@

\begin{lyxcode}
% \centering
Fujian~25.917~118.283

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig003}
\caption{
{ \footnotesize 
{\bf Location coordinates editor.}
} % END: footnotesize
}
\label{fig:003}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig0003}
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:0003}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig03}
\caption{
{ \footnotesize 
{\bf Color chooser.}
} % END: footnotesize
}
\label{fig:03}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig04}
\caption{
{ \footnotesize 
{\bf Message in Terminal}
} % END: footnotesize
}
\label{fig:04}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{H5N1_discrete_vis}
\caption{
{ \footnotesize 
{\bf Screenshot of the Spread output.}
} % END: footnotesize
}
\label{fig:05}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has generated a BEAST log file with
rate indicators as described in Bayesian stochastic search variable
selection (BSSVS) procedure, as described here:

\url{http://beast.bio.ed.ac.uk/BSSVS}

\noindent
The test aims at identifying frequently invoked rates to explain the
diffusion process and visualize them in virtual globe software or
using Spread own map.

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig06}
\caption{
{ \footnotesize 
{\bf Loading the log file.}
} % END: footnotesize
}
\label{fig:06}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig006}
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:006}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{H5N1_bf_vis}
\caption{
{ \footnotesize 
{\bf Generated visualisations.}
} % END: footnotesize
}
\label{fig:07}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.2]{fig08}
\caption{
{ \footnotesize 
{\bf Terminal tab output.}
} % END: footnotesize
}
\label{fig:08}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has set up a BEAST phylogeographic
analysis in continuous space and generated a Maximum Clade Credibility
(MCC) tree using TreeAnnotator. Good tutorial on neccessary steps
can be found here:

\url{http://beast.bio.ed.ac.uk/Continuous_phylogeographic_analysis}

\noindent
This visualisation aims at projecting the MCC tree on the grid of
geographical coordinates, with polygons representating the uncertainty
in location coordinates and considers raccoon rabies diffusion in
north-eastern United States. 
The example tree file used in the presented analysis can be accessed
from here:

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig09}
\caption{
{ \footnotesize 
{\bf Loading tree file.}
} % END: footnotesize
}
\label{fig:09}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{RacRABV_cont_vis-bw}
\caption{
{ \footnotesize 
{\bf Screenshot of the Spread output.}
} % END: footnotesize
}
\label{fig:RacRABV_cont_vis}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has set up a BEAST phylogeographic
analysis in continuous space and generated a trees file and an MCC
tree file using TreeAnnotator. Good tutorial on neccessary steps can
be found here:

\url{http://beast.bio.ed.ac.uk/Continuous_phylogeographic_analysis}

\noindent
This analysis allows one to summarise and visualise full posterior
distribution of the trees obtained in continuous phylogeographic analysis.
To achieve this Spread creates a time line according to the MCC tree
length, slices through each phylogeny at a particular points in time,
imputes the unobserved descendant locations for those time points
and contoures them by creating polygons, a natural representation
of the uncertainty in these inferences. In this template we also visualise
the MCC tree which gave rise to the time slices by drawing it's branches.

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.2]{fig10}
\caption{
{ \footnotesize 
{\bf Choosing analysis type.}
} % END: footnotesize
}
\label{fig:10}
\par\end{centering}
\end{figure}

@@

\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{RacRABV_Time_Slicer_vis}
\caption{
{ \footnotesize 
{\bf Terminal tab output.}
} % END: footnotesize
}
\label{fig:12}
\par\end{centering}
\end{figure}

>>>>

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig01}
\caption{
{ \footnotesize 
{\bf Loading MCC tree file.}
} % END: footnotesize
}
\label{fig:01}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig02}
\caption{
{ \footnotesize 
{\bf Message in Terminal.}
} % END: footnotesize
}
\label{fig:02}
\par\end{centering}
\end{figure}

@@

\begin{lyxcode}
% \centering
% \footnotesize{
Fujian~25.917~118.283

@@

Hunan~27.383~111.517
% }
\end{lyxcode}

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig003}
\caption{
{ \footnotesize 
{\bf Location coordinates editor.}
} % END: footnotesize
}
\label{fig:003}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig0003}
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:0003}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig03}
\caption{
{ \footnotesize 
{\bf Color chooser.}
} % END: footnotesize
}
\label{fig:03}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig04}
\caption{
{ \footnotesize 
{\bf Message in Terminal}
} % END: footnotesize
}
\label{fig:04}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{H5N1_discrete_vis}
\caption{
{ \footnotesize 
{\bf Screenshot of the Spread output.}
} % END: footnotesize
}
\label{fig:05}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has generated a BEAST log file with
rate indicators as described in Bayesian stochastic search variable
selection (BSSVS) procedure, as described at \url{http://beast.bio.ed.ac.uk/BSSVS}.

The test aims at identifying frequently invoked rates to explain the
diffusion process and visualize them in virtual globe software or
using Spread own map.

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig06}
\caption{
{ \footnotesize 
{\bf Loading the log file.}
} % END: footnotesize
}
\label{fig:06}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig006}
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:006}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{H5N1_bf_vis}
\caption{
{ \footnotesize 
{\bf Generated visualisations.}
} % END: footnotesize
}
\label{fig:07}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.2]{fig08}
\caption{
{ \footnotesize 
{\bf Terminal tab output.}
} % END: footnotesize
}
\label{fig:08}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has set up a BEAST phylogeographic
analysis in continuous space and generated a Maximum Clade Credibility
(MCC) tree using TreeAnnotator. Good tutorial on neccessary steps
can be found at \url{http://beast.bio.ed.ac.uk/Continuous_phylogeographic_analysis}.

% \noindent
This visualisation aims at projecting the MCC tree on the grid of
geographical coordinates, with polygons representating the uncertainty
in location coordinates and considers raccoon rabies diffusion in
north-eastern United States. 
The example tree file used in the presented analysis can be accessed
from \url{http://www.kuleuven.be/aidslab/phylogeography/SPREAD_files/RacRABV_cont_0.8_MCC_snyder.tre}.

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{fig09}
\caption{
{ \footnotesize 
{\bf Loading tree file.}
} % END: footnotesize
}
\label{fig:09}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{RacRABV_cont_vis-color}
\caption{
{ \footnotesize 
{\bf Screenshot of the Spread output.}
} % END: footnotesize
}
\label{fig:RacRABV_cont_vis}
\par\end{centering}
\end{figure}

@@

This tutorial assumes the user has set up a BEAST phylogeographic
analysis in continuous space and generated a trees file and an MCC
tree file using TreeAnnotator. Good tutorial on neccessary steps can
be found at \url{http://beast.bio.ed.ac.uk/Continuous_phylogeographic_analysis}.

% \noindent
This analysis allows one to summarise and visualise full posterior
distribution of the trees obtained in continuous phylogeographic analysis.
To achieve this Spread creates a time line according to the MCC tree
length, slices through each phylogeny at a particular points in time,
imputes the unobserved descendant locations for those time points
and contoures them by creating polygons, a natural representation
of the uncertainty in these inferences. In this template we also visualise
the MCC tree which gave rise to the time slices by drawing it's branches.

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.2]{fig10}
\caption{
{ \footnotesize 
{\bf Choosing analysis type.}
} % END: footnotesize
}
\label{fig:10}
\par\end{centering}
\end{figure}

@@

%\begin{figure}[h!]
\begin{figure}[H]
\begin{centering}
\includegraphics[scale=0.25]{RacRABV_Time_Slicer_vis}
\caption{
{ \footnotesize 
{\bf Terminal tab output.}
} % END: footnotesize
}
\label{fig:12}
\par\end{centering}
\end{figure}

### 2015-03-11 appendices/appendixc.tex
cover for new, thicker version

\begin{code}
@Article\{bielejec2014, \\
AUTHOR = \{Bielejec, Filip and Lemey, Philippe and Carvalho, Luiz
and Baele, Guy and Rambaut, Andrew and Suchard, Marc A.\}, \\
TITLE = \{piBUSS: a parallel BEAST/BEAGLE utility for sequence simulation
under complex evolutionary scenarios\}, \\
JOURNAL = \{BMC Bioinformatics\}, \\
VOLUME = \{15\}, \\
YEAR = \{2014\}, \\
NUMBER = \{1\}, \\
PAGES = \{133\} \\
\}
\end{code}

>>>>

% \vspace{-3.5cm}

\begin{code}
\footnotesize{
@Article\{bielejec2014, \\
  AUTHOR = \{Bielejec, Filip and Lemey, Philippe and Carvalho, Luiz
and Baele, Guy and Rambaut, Andrew and Suchard, Marc A.\}, \\
  TITLE = \{piBUSS: a parallel BEAST/BEAGLE utility for sequence simulation
under complex evolutionary scenarios\}, \\
  JOURNAL = \{BMC Bioinformatics\}, \\
  VOLUME = \{15\}, \\
  YEAR = \{2014\}, \\
  NUMBER = \{1\}, \\
  PAGES = \{133\} \\
\}
}
\end{code}

### 2015-03-11 chapters/chapter02.tex
dates, cover, typos
changes for printout
cover for new, thicker version

From the rate matrix $\mathbf{Q}$, a stochastic matrix $\mathbf{P}$ is computed over time $t \geq 0$ via matrix exponentiation $\mathbf{P}(t)=
\exp(t\mathbf{Q})
=\underset{j=0}{\overset{\infty}{\sum}}
(t\mathbf{Q})^{j} / j!$. For an overview of methods to numerically approximate a matrix exponential, we refer to \citet{Moler1978}.
%
Drawing realizations with probabilities defined by $\mathbf{P}$ gives rise to a stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that %\newline
%
$P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n},\ldots, X(t_{0})=i_{0}\right\} =P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n}\right\} $. 
%
In general, one refers to the elements of $\mathbf{P}$ as finite-time transition probabilities between the $K$ discrete state-space elements. Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by

@@

A tutorial which explains how to set up an `epoch model' for analysis in BEAST?BEAGLE can be found in Appendix~\ref{app:spread_tuto}.

@@

%---SHANKARAPPA ANALYSIS RESULTS HPD---%
\begin{figure}[h!]
\centering
% \color{fgcolor}
\includegraphics[scale=0.6]{shankarappa_hpd_new} 
\caption{
{ \footnotesize 
{\bf Estimates of $d_N/d_S$ ratio for within-host HIV analyses.} Vertical lines represent 95\% highest posterior density intervals for the $d_N/d_S$ ratio estimates. 
Parameter $\omega$ is estimated under the homogeneous model, while $\omega_{1}$ and $\omega_{2}$ are obtained using the epoch model.
}% END: footnotesize
}
\label{fig:shank_hpd}
\end{figure}

@@

%---SHANKARAPPA ANALYSIS RESULTS BF---%
\begin{table}[h!]
\centering
% \medskip
% \begin{minipage}{\textwidth} 
% \footnotesize{
\begin{tabular}%{ccc}
{@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
\textbf{Patient} & \textbf{Posterior probability} & \textbf{log Bayes factor}\tabularnewline
\hline 
patient 1 & $0.999$ & $7.418$\tabularnewline
patient 2 & >$0.999$ & $9.602$\tabularnewline
patient 3 & $0.898$ & $2.174$\tabularnewline
patient 5 & $0.430$ & $-0.282$\tabularnewline
patient 6 & >$0.999$ & $9.210$\tabularnewline
patient 7 & >$0.999$ & $8.112$\tabularnewline
patient 8 & $0.933$ & $2.627$\tabularnewline
patient 9 & $0.895$ & $2.142$\tabularnewline
\hline 
\textbf{Joint evidence:} & $0.894$ & $2.14$ \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Bayes factor test for decreased selection after progression.} We report the posterior probability that $\omega_{1} < \omega_{2}$ and the corresponding Bayes factor against the alternative that $\omega_{1} \ge \omega_{2}$.
}% END: footnotesize
}
\label{tab:shank_bf}
% \end{minipage}
\end{table}

Despite the observation that the Bayesian credible intervals for patient 1, 2, 6 and 7 estimates do not overlap, this does not provide a formal test to evaluate their differences.
Therefore, we conduct a Bayes factor (BF) test \citep{Suchard2005} that expresses the posterior odds over the prior odds that $\omega_{1} < \omega_{2}$ for the individual analyses of each patient.
% To determine the posterior odds, we note that the MCMC sample average of an indicator function that the parameter values fall within one competing model space converges to the posterior probability of that model.
% \myedit{sveighteen}{
To determine the posterior odds for the hypothesis that $\omega_{1} < \omega_{2}$, we note that the MCMC sample average of an indicator function that keeps track of samples for which $\omega_{1} < \omega_{2}$ converges to the posterior probability of that hypothesis.
% }
The prior odds in our case is simply 1.
The log Bayes factors listed in Table \ref{tab:shank_bf} suggest generally strong evidence for a declining selective pressure after progression, with one notable exception for patient 5.
We also provide a Bayes factor that summarizes the joint evidence for $\omega_{1} < \omega_{2}$, which suggest an overall support in favor of the immune relaxation hypothesis (log BF = 2.14), in accordance with previous findings suggesting a general decrease in non-synonymous divergence at late-stage infection \citep{williamson2005, Lemey2007}.

@@

%---FLU DATA RESULTS MAP---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.5]{flu2_epoch} 
\caption{
{ \footnotesize 
{\bf A two-epoch phylogeographic model applied to seasonal influenza H3N2.} 
A. Maximum clade credibility (MCC) tree with branches colored according to modal discrete location states at each node. The grey time intervals represent the epoch model with a single discrete rate matrix shared across northern hemisphere spring and summer (light grey) time intervals and another rate matrix shared across the northern hemisphere autumn and winter (dark grey) time intervals. B. Diffusion rates supported by a Bayes factor $>20$ for spring and summer epoch intervals. The width of the arrows reflects the magnitude of the Bayes factor support. C. Diffusion rates supported by a Bayes factor $>20$ for autumn and winter epoch intervals.
}% END: footnotesize 
}
\label{fig:2EpochFlu}
\end{figure}

@@

%---FLU DATA RESULTS PS SS---%
% \afterpage{\clearpage
\begin{table}[h!]
\begin{minipage}{\textwidth} 
\centering
% \footnotesize{
\begin{tabular}{ccc}
\hline 
\multirow{2}{*}{\textbf{Model}} & \multicolumn{2}{c}{\textbf{Marginal likelihood}}\tabularnewline
 & PS\footnote{Path Sampling} & SS\footnote{Stepping Stone Sampling}\tabularnewline
\hline 
homogeneous & -827.29 & -825.07\tabularnewline
7-epoch & -806.36  & -803.40 \tabularnewline
14-epoch & -798.77  & -795.63 \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Marginal likelihood estimates.} Comparison in terms of model fit between a homogeneous model, an epoch model with time discretized into $S=7$ epochs alternating between 2 different rate matrices and an epoch model with time discretized into $S=14$ epochs, alternating between 4 separate rate matrices.
} % END: footnotesize
}
\label{tab:flu_ps}
\end{minipage}
\end{table}
% \clearpage }% END: afterpage

@@

Although the hypothesis we test here does not require investigating site-specific selection patterns, we note that codon models that accommodate different categories of sites also have interesting applications in the epoch framework.
This has been demonstrated by \citet{Goode2008}, who employed a particular codon parameterization to allow for sites to switch among a neutrally evolving class, a class of negatively selected sites and a class of positively selected sites (cfr. M2, \citet{NY98}) before and after the start of HIV-1 antiretroviral therapy.
Using this approach, the authors showed that a considerable number of sites, which are identified as neutral under a time-homogenous model, are under some form of selection in one of the two epoch time periods, implying that neutrality may be context-dependent.
Codon models that accommodate different categories of sites have not yet been implemented in BEAST, at least partly because they are computationally prohibitive.
Incorporating such models in an epoch approach would contribute even more to the computational burden.
We hope however that the parallel implementations we pursue here may still allow practitioners to explore such extensions in future research. %

Our second application exemplifies  the use of epoch time-discretization in phylogeographic inferences that consider sampling locations as discrete traits.
In particular, we demonstrate how epoch modeling can capture seasonality in human influenza A H3N2 circulation dynamics, without the need to complicate location traits with sampling time.
Based on a data set previously analyzed by \citet{Bahl2011}, we infer different epidemiological connections between the northern hemisphere spring-summer epoch and the autumn-winter epoch.

>>>>

From the rate matrix $\mathbf{Q}$, a stochastic matrix $\mathbf{P}$ is computed over time $t \geq 0$ via matrix exponentiation $\mathbf{P}(t)=
\exp(t\mathbf{Q})
=\underset{j=0}{\overset{\infty}{\sum}}
(t\mathbf{Q})^{j} / j!$. For an overview of methods to numerically approximate a matrix exponential, we refer to \citet{Moler1978}.
%
Drawing realizations with probabilities defined by $\mathbf{P}$ gives rise to a stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that %\newline

\begin{align}
& P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n},\ldots, X(t_{0})=i_{0}\right\} \nonumber \\
&= P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n}\right\}. 
\end{align}

In general, one refers to the elements of $\mathbf{P}$ as finite-time transition probabilities between the $K$ discrete state-space elements. Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by

@@

A tutorial which explains how to set up an `epoch model' for analysis in BEAST can be found in Appendix~\ref{app:spread_tuto}.

@@

%---SHANKARAPPA ANALYSIS RESULTS HPD---%
\begin{figure}[H]
\centering
% \color{fgcolor}
\includegraphics[scale=0.6]{shankarappa_hpd_new} 
\caption{
{ \footnotesize 
{\bf Estimates of $d_N/d_S$ ratio for within-host HIV analyses.} Vertical lines represent 95\% highest posterior density intervals for the $d_N/d_S$ ratio estimates. 
Parameter $\omega$ is estimated under the homogeneous model, while $\omega_{1}$ and $\omega_{2}$ are obtained using the epoch model.
}% END: footnotesize
}
\label{fig:shank_hpd}
\end{figure}

@@

Despite the observation that the Bayesian credible intervals for patient 1, 2, 6 and 7 estimates do not overlap, this does not provide a formal test to evaluate their differences.
Therefore, we conduct a Bayes factor (BF) test \citep{Suchard2005} that expresses the posterior odds over the prior odds that $\omega_{1} < \omega_{2}$ for the individual analyses of each patient.
% To determine the posterior odds, we note that the MCMC sample average of an indicator function that the parameter values fall within one competing model space converges to the posterior probability of that model.
% \myedit{sveighteen}{
To determine the posterior odds for the hypothesis that $\omega_{1} < \omega_{2}$, we note that the MCMC sample average of an indicator function that keeps track of samples for which $\omega_{1} < \omega_{2}$ converges to the posterior probability of that hypothesis.
% }
The prior odds in our case is simply 1.
The log Bayes factors listed in Table \ref{tab:shank_bf} suggest generally strong evidence for a declining selective pressure after progression, with one notable exception for patient 5.

%---SHANKARAPPA ANALYSIS RESULTS BF---%
\begin{table}[H]
\centering
% \medskip
% \begin{minipage}{\textwidth} 
\footnotesize{
\begin{tabular}%{ccc}
{@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
\textbf{Patient} & \textbf{Posterior probability} & \textbf{log Bayes factor}\tabularnewline
\hline 
patient 1 & $0.999$ & $7.418$\tabularnewline
patient 2 & >$0.999$ & $9.602$\tabularnewline
patient 3 & $0.898$ & $2.174$\tabularnewline
patient 5 & $0.430$ & $-0.282$\tabularnewline
patient 6 & >$0.999$ & $9.210$\tabularnewline
patient 7 & >$0.999$ & $8.112$\tabularnewline
patient 8 & $0.933$ & $2.627$\tabularnewline
patient 9 & $0.895$ & $2.142$\tabularnewline
\hline 
\textbf{Joint evidence:} & $0.894$ & $2.14$ \tabularnewline
\end{tabular}
} % END: footnotesize
\caption{
{ \footnotesize 
{\bf Bayes factor test for decreased selection after progression.} We report the posterior probability that $\omega_{1} < \omega_{2}$ and the corresponding Bayes factor against the alternative that $\omega_{1} \ge \omega_{2}$.
}% END: footnotesize
}
\label{tab:shank_bf}
% \end{minipage}
\end{table}

We also provide a Bayes factor that summarizes the joint evidence for $\omega_{1} < \omega_{2}$, which suggest an overall support in favor of the immune relaxation hypothesis (log BF = 2.14), in accordance with previous findings suggesting a general decrease in non-synonymous divergence at late-stage infection \citep{williamson2005, Lemey2007}.

@@

%---FLU DATA RESULTS MAP---%
\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{flu2_epoch} 
\caption{
{ \footnotesize 
{\bf A two-epoch phylogeographic model applied to seasonal influenza H3N2.} 
A. Maximum clade credibility (MCC) tree with branches colored according to modal discrete location states at each node. The grey time intervals represent the epoch model with a single discrete rate matrix shared across northern hemisphere spring and summer (light grey) time intervals and another rate matrix shared across the northern hemisphere autumn and winter (dark grey) time intervals. B. Diffusion rates supported by a Bayes factor $>20$ for spring and summer epoch intervals. The width of the arrows reflects the magnitude of the Bayes factor support. C. Diffusion rates supported by a Bayes factor $>20$ for autumn and winter epoch intervals.
}% END: footnotesize 
}
\label{fig:2EpochFlu}
\end{figure}

@@

%---FLU DATA RESULTS PS SS---%
% \afterpage{\clearpage
\begin{table}[h!]
\begin{minipage}{\textwidth} 
\centering
\footnotesize{
\begin{tabular}{ccc}
\hline 
\multirow{2}{*}{\textbf{Model}} & \multicolumn{2}{c}{\textbf{Marginal likelihood}}\tabularnewline
 & PS\footnote{Path Sampling} & SS\footnote{Stepping Stone Sampling}\tabularnewline
\hline 
homogeneous & -827.29 & -825.07\tabularnewline
7-epoch & -806.36  & -803.40 \tabularnewline
14-epoch & -798.77  & -795.63 \tabularnewline
\end{tabular}
} % END: footnotesize
\caption{
{ \footnotesize 
{\bf Marginal likelihood estimates.} Comparison in terms of model fit between a homogeneous model, an epoch model with time discretized into $S=7$ epochs alternating between 2 different rate matrices and an epoch model with time discretized into $S=14$ epochs, alternating between 4 separate rate matrices.
} % END: footnotesize
}
\label{tab:flu_ps}
\end{minipage}
\end{table}
% \clearpage }% END: afterpage

@@

Although the hypothesis we test here does not require investigating site-specific selection patterns, we note that codon models that accommodate different categories of sites also have interesting applications in the epoch framework.
This has been demonstrated by \citet{Goode2008}, who employed a particular codon parameterization to allow for sites to switch among a neutrally evolving class, a class of negatively selected sites and a class of positively selected sites (cfr. M2, \citet{NY98}) before and after the start of HIV-1 antiretroviral therapy.
Using this approach, the authors showed that a considerable number of sites, which are identified as neutral under a time-homogenous model, are under some form of selection in one of the two epoch time periods, implying that neutrality may be context-dependent.
Codon models that accommodate different categories of sites have not yet been implemented in BEAST, at least partly because they are computationally prohibitive.
Incorporating such models in an epoch approach would contribute even more to the computational burden.
We hope however that the parallel implementations we pursue here may still allow practitioners to explore such extensions in future research.

% \nohyphens{
Our second application exemplifies  the use of epoch time discretization in phylogeographic inferences that consider sampling locations as discrete traits.
In particular, we demonstrate how epoch modeling can capture seasonality in human influenza A H3N2 circulation dynamics, without the need to complicate location traits with sampling time.
Based on a data set previously analyzed by \citet{Bahl2011}, we infer different epidemiological connections between the northern hemisphere spring-summer epoch and the autumn-winter epoch.
% }

### 2015-03-11 chapters/chapter04.tex
cover for new, thicker version

The core of {\bussname} consists of a recursive tree-traversal that is independent of the BEAST inference machinery. 
The algorithm simulates discrete state realizations by visiting the tree nodes in pre-order fashion, i.e., parental nodes are visited before child nodes.
When a child node is visited, {\bussname} samples its state from the conditional probabilities of changing to state $j$ given state $i$ at the parental node.
For a branch length $t$ and clock rate $r$, the finite-time transition probability matrix $\mathbf{P}\left(r \times t\right)$ is calculated through the eigen-decomposition of the infinitesimal rate matrix $\mathbf{Q}$ along that branch. 
For a review of methods to numerically approximate a matrix exponential, we refer to \cite{Moler1978}.

>>>>

The core of {\bussname} consists of a recursive tree-traversal that is independent of the BEAST inference machinery. 
The algorithm simulates discrete state realizations by visiting the tree nodes in pre-order fashion, i.e., parental nodes are visited before child nodes.
When a child node is visited, {\bussname} samples its state from the conditional probabilities of changing to state $j$ given state $i$ at the parental node.
For a branch length $t$ and clock rate $r$, the finite-time transition probability matrix $\mathbf{P}\left(r\times t\right)$ is calculated through the eigen-decomposition of the infinitesimal rate matrix $\mathbf{Q}$ along that branch. 
For a review of methods to numerically approximate a matrix exponential, we refer to \cite{Moler1978}.
